HaPy \\ Haskell for Python

### 2011-12-17 paper.tex
Some last-minute updates to the paper (got namerefs working)

\section{Use}
Our goal with HaPy is to allow the user to call arbitrary Haskell code while remaining committed to ``Pythonic'' syntax. In particular, importing a Haskell module into a Python program is easy with HaPy: simply add the line
\begin{minted}{python}
from HaPy import HsModule
\end{minted}
or
\begin{minted}{python}
import HaPy.HsModule
\end{minted}
to the head of any Python script. This will dynamically load the specified Haskell module, so that any Haskell function defined in HsModule can be accessed and executed by the Python script, as though it were code defined in another Python module. For instance, if HsModule defines a function \verb!sum! which takes two real numbers and returns their sum, the result can be printed in Python with
\begin{minted}{python}
print HsModule.sum(first, second)
\end{minted}
Installing HaPy is similarly intuitive, but with the caveat that all libraries called by \verb!HaPy.hs! must be dynamically-linkable. This means that all of the packages that \verb!HaPy.hs! imports must be reinstalled with Cabal, using the \verb!--enable-shared! flag. \footnote{The requirement that these packages be dynamically-linkable stems from the fact that \texttt{ctypes} must be able to dynamically link to \texttt{HaPy.hs} (see \nameref{simple_install}) } \\
However, Haskell packages that are used by the user's Haskell code need not be dynamically-linkable. Indeed, on the Haskell side, no special support is needed by the user; any standard Haskell code will run with HaPy (modulo the incompleteness of our type checker). All the issues of marshalling data and passing data to and from the Haskell foreign function interface are managed by the code in \verb!HaPy.hs!. This has the added benefit of enabling the user to call any code defined in the Haskell standard libraries from Python. 
% - Meshing python and haskell semantics

@@

\subsection{Haskell Types in Python}
One of the main challenges in HaPy is the representation of Haskell data types in Python.  Some primitive Haskell types (currently \verb!Bool!, \verb!Int!, \verb!Double!, and \verb!String!) are marshalled into the equivalent Python type.  All other Haskell types are returned as opaque references: they can be passed to Haskell functions, but cannot be modified or inspected by other Python code.
Use of Haskell functions is trivial: they can be called just like other Python functions.  Arguments can be any of the supported primitive types, which will be marshalled, or Haskell object references, which are passed through.  The return value will be automatically converted into a Python type if possible; otherwise, it will remain as an opaque Haskell reference.
The Haskell functions do not behave exactly like other Python functions: they support currying.  When they are called with too few arguments, a partially-applied function object is returned.  This does not invalidate the original function - all functions and partially-applied functions can be called as many times as desired.  We chose to do this because it naturally followed from the implementation (see \nameref{imp_currying}) and nicely mirrors the behavior of actual Haskell functions.  Further, this functionality can be ignored if not desired.  (The user only loses immediate too-few-argument errors.)

@@

\subsection{Function Currying}
\label{sec:imp_currying}

@@

\subsection{Simplified Install}
\label{sec:simple_install}
While developing HaPy, we encountered a serious bug relating to dynamic linking. In 64-bit environments, this results segfaults when using the Haskell \verb!System.Plugins! library. \\
Further, the Python \verb!ctypes! module dynamically links to our binary. Unfortunately, this requires that we build our binary with the \verb!-dynamic! flag, which causes GHC to link all of the libraries HaPy uses dynamically. This means that users of HaPy must use Cabal to reinstall several packages with the \verb!--enable-shared! option. We should be able to statically link all of our libraries, so that we can simply distribute a self-contained binary. Furthermore, we speculate that this might solve our 64-bit bug.

>>>>

\section{Use}
Our goal with HaPy is to allow the user to call arbitrary Haskell code while remaining committed to ``Pythonic'' syntax. In particular, importing a Haskell module into a Python program is easy with HaPy: simply add the line
\begin{minted}{python}
from HaPy import HsModule
\end{minted}
or
\begin{minted}{python}
import HaPy.HsModule
\end{minted}
to the head of any Python script. This will dynamically load the specified Haskell module, so that any Haskell function defined in HsModule can be accessed and executed by the Python script, as though it were code defined in another Python module. For instance, if HsModule defines a function \verb!sum! which takes two real numbers and returns their sum, the result can be printed in Python with
\begin{minted}{python}
print HsModule.sum(first, second)
\end{minted}
Installing HaPy is similarly intuitive, but with the caveat that all libraries called by \verb!HaPy.hs! must be dynamically-linkable. This means that all of the packages that \verb!HaPy.hs! imports must be reinstalled with Cabal, using the \verb!--enable-shared! flag. \footnote{The requirement that these packages be dynamically-linkable stems from the fact that \texttt{ctypes} must be able to dynamically link to \texttt{HaPy.hs} (see \nameref{sec:simple_install}) } \\
However, Haskell packages that are used by the user's Haskell code need not be dynamically-linkable. Indeed, on the Haskell side, no special support is needed by the user; any standard Haskell code will run with HaPy (modulo the incompleteness of our type checker). All the issues of marshalling data and passing data to and from the Haskell foreign function interface are managed by the code in \verb!HaPy.hs!. This has the added benefit of enabling the user to call any code defined in the Haskell standard libraries from Python. 
% - Meshing python and haskell semantics

@@

\subsection{Haskell Types in Python}
One of the main challenges in HaPy is the representation of Haskell data types in Python.  Some primitive Haskell types (currently \verb!Bool!, \verb!Int!, \verb!Double!, and \verb!String!) are marshalled into the equivalent Python type.  All other Haskell types are returned as opaque references: they can be passed to Haskell functions, but cannot be modified or inspected by other Python code.
Use of Haskell functions is trivial: they can be called just like other Python functions.  Arguments can be any of the supported primitive types, which will be marshalled, or Haskell object references, which are passed through.  The return value will be automatically converted into a Python type if possible; otherwise, it will remain as an opaque Haskell reference.
The Haskell functions do not behave exactly like other Python functions: they support currying.  When they are called with too few arguments, a partially-applied function object is returned.  This does not invalidate the original function - all functions and partially-applied functions can be called as many times as desired.  We chose to do this because it naturally followed from the implementation (see \nameref{sec:imp_currying}) and nicely mirrors the behavior of actual Haskell functions.  Further, this functionality can be ignored if not desired.  (The user only loses immediate too-few-argument errors.)

@@

\subsection{Function Currying}\label{sec:imp_currying}

@@

\subsection{Simplified Install}\label{sec:simple_install}
While developing HaPy, we encountered a serious bug relating to dynamic linking. In 64-bit environments, this results segfaults when using the Haskell \verb!System.Plugins! library. \\
Further, the Python \verb!ctypes! module dynamically links to our binary. Unfortunately, this requires that we build our binary with the \verb!-dynamic! flag, which causes GHC to link all of the libraries HaPy uses dynamically. This means that users of HaPy must use Cabal to reinstall several packages with the \verb!--enable-shared! option. We should be able to statically link all of our libraries, so that we can simply distribute a self-contained binary. Furthermore, we speculate that this might solve our 64-bit bug.
